Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-08-11 Manuscript/MPESubmission/Revision/response.tex

% NC: Needs a little more explanation, and a line about what analysis you did in appendix b. So something along the lines of "We compare best trees to missing data trees, but could also compare true trees to the missing data trees. We do not do this because... However, results of analyses comparing the two can be found in appendix B. This makes little difference to our overall results." That might need to be better integrated with the paragraph(s) before.
% TG: changed to:
"We compare ``best'' trees to ``missing data'' trees but could also compare ``true'' trees to the ``missing data'' trees.
In practice, the difference between the ``best'' trees and the ``missing data'' trees represents the effect of our missing data parameters and of the phylogenetic methods used to infer the ``missing data'' trees.
The difference between the ``true'' and the ``missing data'' tree, however, represents the effect of our parameters used to generate the ``true'' tree and the algorithms used to generated the ``complete'' matrix as well as the effect of our missing data parameters and the phylogenetic methods used.
Because the main aim of this study, is to look at the effect the missing data parameters on topological recovery, we chose to represent only the comparisons between the ``best'' trees to ``missing data'' trees.
The results of the comparisons of the ``true'' and the ``missing data'' tree are available in appendix B.
Note that these results makes little difference to our overall results." line 276-280 @@@.

See also appendix B section "Differences between the ``true'' and the ``best'' trees." and reviewer 2's comment 1 below.

@@

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\caption{Distribution overlap between Maximum Likelihood (ML) and Bayesian consensus tree (BI) across all missing data parameters combinations (pooled) and every single missing data parameter individually ($M_L$, $M_F$, $N_C$)}
\centering
\begin{tabular}{|l|c|c|c|c|c|c|c|}
  \hline
 Comparison &  Metric & Min. & 1st Qu. & Median & Mean & 3rd Qu. & Max. \\ 
  \hline
    ML \textit{vs.} BI (pooled) & $RF$ & \textbf{0.00} & \textbf{0.00} & 0.10 & 0.20 & 0.32 & \textbf{1.00} \\ 
    ML \textit{vs.} BI ($M_L$)  & $RF$ & 0.30          & 0.31          & 0.69 & 0.61 & 0.77 & \textbf{1.00} \\ 
    ML \textit{vs.} BI ($M_F$)  & $RF$ & \textbf{0.00} & 0.25          & 0.48 & 0.50 & 0.76 & \textbf{1.00} \\ 
    ML \textit{vs.} BI ($N_C$)  & $RF$ & \textbf{0.03} & 0.32          & 0.66 & 0.55 & 0.75 & \textbf{1.00} \\
    ML \textit{vs.} BI (pooled) & $Tr$ & 0.34          & 0.49          & 0.61 & 0.62 & 0.75 & \textbf{1.00} \\
    ML \textit{vs.} BI ($M_L$)  & $Tr$ & 0.79          & 0.81          & 0.84 & 0.86 & 0.85 & \textbf{1.00} \\  
    ML \textit{vs.} BI ($N_C$)  & $Tr$ & 0.51          & 0.69          & 0.80 & 0.76 & 0.80 & \textbf{1.00} \\
    ML \textit{vs.} BI ($M_F$)  & $Tr$ & 0.38          & 0.69          & 0.75 & 0.72 & 0.80 & \textbf{1.00} \\  
   \hline
\end{tabular}
\label{Tab_Results-Difference_methods}
%\end{adjustwidth}
\end{table}

2) We followed the reviewer's suggestion and added 6 extra tables in the Appendix C (Table C8 to C13) containing the marginal probability of each missing data parameter individually and each tree inference method.
We considered running the same type of analysis for the combination of the parameter, however, this involved a table of 5*126 cells which was highly impractical to read.
% TG: Not sure if the last bit is necessary...

@@

% -----------------
% Specific Comment 59
% -----------------
\item{\textcolor{blue}{\textbf{Appendix C:} Tables 1-3 repeats 'of the'. Should label the first two rows as 'Best maximum likelihood tree' and 'Best Bayesian inference tree'}}

We removed the repetition in the caption.
If we understood the second point correctly, the reviewer suggests we add ``\textit{best}'' to the two first rows of the tables, and by ``\textit{best}'', the reviewer means ``optimal'' (as opposed to the ``best'' tree in our protocol that refers to any tree inferred by any method from the ``complete matrix''.
We don't think it is necessary to follow this suggestion as the Maximum Likelihood tree is by definition the tree with the highest likelihood, and the Bayesian consensus is the consensus tree of the Bayesian inference trees.
So adding the word ``\textit{best}'' here is not entirely accurate.

% -----------------
% Specific Comment 60
% -----------------
\item{\textcolor{blue}{\textbf{Appendix D:} I find the online code repository thorough, but the repository is loaded with many files, especially for graphing. If it is possible to condense these files it may be easier for a reader to duplicate the study.}}

We updated the README page of the GitHub repository linking to the various R code to make it clearer what files are needed.
We also now provide a version of the analysis in R run thoughout the manuscript as a step by step tutorial for a subset of data.
See \url{https://github.com/TGuillerme/Total_Evidence_Method-Missing_data/blob/master/Analysis/Analysis_Demo.R} and Appendix D.
% NC: Maybe add something like this if this is your intention. It's probably worth it. No need to keep lots of un-necessary files. TG: Done and agree
We will also clean up this repository once the paper has been accepted.
At present it is functioning both as our working repository for code as well as the repository for this paper.

@@

\end{enumerate}

\section{Reviewer 2:}
Note that we split the second comment from reviewer 2 in three different comments (comments 2,3 and 4 below) to improve the readability of our response to the reviewers.

% -----------------
% Comment 1
% -----------------
\begin{enumerate}
\item{\textcolor{blue}{\textbf{Compare the trees to the "true" tree:} I don't see how the "best" tree approach is helpful in this case. It adds an additional layer of complexity that can't be accounted for via simulation. The point of simulations is to compare empirical realities to the generating process, to see how much results diverge. I think the most helpful and straightforward thing to do is eliminate the "best" tree part of the MS, and compare the inferred trees directly to the real topology to quantify variance.}}

A similar point/suggestion was also raised by reviewer 1 and we ran the suggested extra analysis (see Reviewer 1 General comment number 6 and specific comment number 25).
We understand the concerns of the reviewer but, we do not agree that comparing the "missing-data" trees to the "best" trees adds a layer of complexity to our simulations.
In practice, the ``true'' tree can be perceived as a ``random'' seed to generate the ``complete'' matrix.
The ``true'' tree is simply a birth-death tree with $\lambda$ $>$ $\mu$ and a set number of living and fossil taxa (see details in the manuscript section "2.1 Generating the matrix").
All these parameters, were only used to create a realistic ``complete'' matrix in order to make our results relevant to biologists.
The aim of this study was to assess how good will a topology inferred from a matrix with missing data (which is likely in biology) compared to the ones inferred form the ``complete'' matrix (the ``best'' tree).

% TG: maybe add a sentence in the manuscript?

% NC: This argument [TG: the one commented out below] makes no sense. The point the reviewer is making is that we go:
% True tree > Best tree > comparisons when we could just go True tree > comparisons. So our approach *is* more complex than their suggestion. Instead the justification here is the same as used for reviewer one - that in real life scenarios we never know the true tree and we want it to be as close to real life as possible. BUT that we have tested their idea and find it makes no difference but all the results are in the appendix.
% TG: you're right, after reflection, I think we should have never mentioned the ``true'' tree and just say in the methods that we used a random Birth-Death tree with such and such parameters to generate the matrix. It's kind of too late to backup by now but the argument was just kind of saying: "we don't care if we can recover the "true" tree because we don't give a flip on how the algorithms perform at making a matrix". It's all random anyway...
% TG : I've commented out the non-sense justification below and wrote something shorter above (and referring to reviewer's 1 comment).

% Because we use a simulation approach to test this problem, we have to start from a random starting point (the "true" tree) and then do the following steps: (1) generate and remove the data (2) infer the topologies and (3) compare the topologies.
% Note that in this schematic representation, the "true" tree is the step (0) of our analysis and involves a series of algorithms and software (i.e. seqGen\{phyclust\}, rTraitDisc\{ape\}, RAxML, MrBayes).
% To compare the "true" tree's topology to the "missing-data" topologies will be equivalent of exploring the effect of our three parameters on recovering topology through 3 layers (1- generate the data; 2- infer the topologies; and 3- compare the topologies).
% By comparing the "best" tree's topology to the "missing-data" ones, we explore the effect of our three parameters through only 2 layers (2- infer the topologies; and 3- compare the topologies).
% Note that we explore the effect of these 2 layers by using two different tree inference methods (effectively 4 since we also use the Maximum Likelihood bootstraps trees and the Bayesian posterior trees) and 2 different Normalised topological metrics (Robinson-Foulds and Triplets).
% Using as main results the comparisons between the "true" and the "missing-data" trees would also need the exploration of the effect of the cascade of algorithms and softwares (maybe by modifying the algorithms used to generate the "true" tree and the matrices?).

% However, we have added analyses with comparisons between the "true" tree and the "missing-data" trees in the Appendix C (Figures C6 and C7).
% It is interesting to note that the algorithms and software as well as probably the size of our matrices are actually not ideal for the recovery of the "true" tree is rather low and scales down the our results (see figure @).% NC: HUH????
% However, even though these comparisons are worse than comparing the best tree to the missing data trees, it is encouraging to see that the patterns in the results are conserved (i.e. Bayesian consensus trees $>$ Maximum Likelihood trees $>$ both bootstraps trees and Bayesian posterior trees).

### 2015-08-11 Manuscript/MPESubmission/Revision/TEM_revised_manuscript.tex
Edited response

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework }\citep[\hl{i.e. using both molecular and morphological data; e.g.}][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
\hl{These studies have assessed the effect of missing data on topology either:(1) by comparing a dataset with missing data to sub-sets without missing data }\citep{Wiens01102005}\hl{; or (2) by removing both molecular and some morphological data from living taxa (``artificial fossil'') }\citep{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Both types of approaches have shown that missing data \hl{are} not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.
However, these studies were limited by their empirical-only approach and the use of only living taxa }\citep{Wiens01102005}\hl{ or the scheme of generating missing data }\citep[\hl{i.e. using the patterns from the fossil record;}][]{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Besides, the studies using fossil data, mainly on the palaeontological aspect of the question }\citep[\hl{i.e. the effect of missing data in fossils in Total Evidence matrices;}][]{manosphylogeny2007,pattinsonphylogeny2014}.
%Former revised version
% \hl{Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework \protect\citep[\hl{i.e. using combined molecular and morphological data}; e.g.][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.}
% \hl{For example, \protect\cite{Wiens01102005} }\hl{only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.}
% \hl{Furthermore, \protect\cite{pattinsonphylogeny2014} }\hl{assess the effect of removing both molecular and morphological data (``artificial extinction'') on the topology.}
% \hl{As for morphological and molecular data separately, these studies shows that missing data are not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.}
% \hl{These studies, however, are limited by their empirical-only approach \protect\citep{Wiens01102005} }\hl{and the scheme of generating missing data \protect\citep{manosphylogeny2007,pattinsonphylogeny2014}.}
% \hl{Also both studies focus on the palaeontological aspect of the question (i.e. the fossil data).}

>>>>

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep[\hl{as long as enough data overlaps in each case, and missing data are not phylogeneticaly biased;}][]{wiensmissing2003,Wiens01102005,wiensmissing2006,wiensmissing2008,lemmonthe2009,Sanderson22072011,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa.
This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices.
\hl{Until now, few attempts have been made to study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework }\citep[\hl{i.e. using both molecular and morphological data; e.g.}][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.
\hl{These studies have assessed the effect of missing data on topology either: (1) by comparing a dataset with missing data to sub-sets without missing data }\citep{Wiens01102005}\hl{; or (2) by removing both molecular and some morphological data from living taxa (``artificial fossil'') }\citep{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Both types of approaches have shown that missing data \hl{are} not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.
However, these studies were limited by their empirical-only approach and the use of only living taxa }\citep{Wiens01102005}\hl{ or the scheme of generating missing data }\citep[\hl{i.e. using the patterns from the fossil record;}][]{manosphylogeny2007,pattinsonphylogeny2014}.
\hl{Besides, the studies using fossil data, mainly on the palaeontological aspect of the question }\citep[\hl{i.e. the effect of missing data in fossils in Total Evidence matrices;}][]{manosphylogeny2007,pattinsonphylogeny2014}.
%Former revised version
% \hl{Until now, only few attempts have been made to study study the impact of this missing data issue on phylogenetic inference in a Total Evidence framework \protect\citep[\hl{i.e. using combined molecular and morphological data}; e.g.][]{Wiens01102005,manosphylogeny2007,pattinsonphylogeny2014}.}
% \hl{For example, \protect\cite{Wiens01102005} }\hl{only assess the effect of missing data by comparing the node support between the complete dataset that contains missing data and sub-datasets with no missing data.}
% \hl{Furthermore, \protect\cite{pattinsonphylogeny2014} }\hl{assess the effect of removing both molecular and morphological data (``artificial extinction'') on the topology.}
% \hl{As for morphological and molecular data separately, these studies shows that missing data are not a major problem and should not be an obstacle to combine both living and fossil species in the same phylogenies.}
% \hl{These studies, however, are limited by their empirical-only approach \protect\citep{Wiens01102005} }\hl{and the scheme of generating missing data \protect\citep{manosphylogeny2007,pattinsonphylogeny2014}.}
% \hl{Also both studies focus on the palaeontological aspect of the question (i.e. the fossil data).}
